Supplemental materials for ``An item response theory analysis of the Matrix Reasoning Item Bank (MaRs-IB)''

### 2022-05-09 manuscript/main.tex
checkpoint manuscript before edits

\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[colorinlistoftodos]{todonotes}

@@

\title{An item response theory analysis of the Matrix Reasoning Item Bank (MaRs-IB)}
\shorttitle{IRT analysis of MaRs-IB}
\author{Samuel Zorowitz$^1$, Gabriele Chierchia(?), Sarah-Jayne Blakemore(?), Nathaniel D. Daw$^{1,2}$}
\affiliation{$^1$Princeton Neuroscience Institute, Princeton University, USA\\$^2$Department of Psychology, Princeton University, USA}

@@

Matrix reasoning tasks are among the most widely used measures of cognitive ability in the behavioral sciences. Much of their popularity undoubtedly reflects their versatility in use. Matrix reasoning tasks are strong (albeit impure) indicators of general intelligence \citep{gignac2015raven} and working memory capacity \citep{kane2004generality, unsworth2005working}. They are predictive of important real-world outcomes such as childhood academic performance \citep{roth2015intelligence} and performance on college entrance exams \citep{frey2004scholastic, koenig2008act}. In low-stakes testing settings, matrix reasoning tasks can additionally function as measures of motivation, willingness to expend mental effort, and other aspects of personality \citep{duckworth2011role, gignac2019maximum}. Performance on matrix reasoning tasks can also serve as an instrumental variable in studies of psychiatric populations; that is, it can be used to account for general disruptions to cognitive ability where specific domains of cognition are primarily of interest \citep{gillan2016characterizing, rouault2018psychiatric, moutoussis2021decision}.

Unfortunately there are a number of impediments to using matrix reasoning tasks as part of behavioral research. One such challenge is the problem of copyright. The most prominent examples of matrix reasoning tasks, such as Raven's progressive matrices \citep{raven2003raven} and the WAIS matrix subtest \citep{wechsler1999wechsler}, are not free to use and have legal restrictions against being computerized. A second challenge is that, across all of the matrix reasoning tests in the public domain, there are relatively few unique items available. The Hagen matrices test \citep{heydasch2014hagen} and ICAR matrix reasoning test \citep{condon2014international}, for example, have only 20 and 16 items respectively. The availability of only a limited number of items raises the possibility of repeat exposure effects, which threaten the validity of these measures  \citep{ng1974applicability, bors2003effect}. This problem is exacerbated in the current age of online experiments, where multiple groups of researchers may be unintentionally administering the same test to the same participants recruited from the same online labor platforms.

Noting these challenges, \cite{chierchia2019matrix} developed and made publicly available the matrix reasoning item bank (MaRs-IB), a collection of 80 unique matrix reasoning puzzles. Each puzzle in the MaRs-IB consists of a 3x3 matrix containing abstract shapes in eight out of nine cells (Figure \ref{fig:fig00}). To solve a puzzle, participants must deduce which of four response options correctly completes the matrix based on the pattern of changing relations among the shapes across cells. The puzzles in the MaRs-IB were designed to span a considerable range of complexity, both in the number of unique shapes and changing relations in and across cells. To further increase the reusability of the MaRs-IB and minimize practice effects, each puzzle (hereafter referred to as a template item) has six clones. An item's clones share identical structural properties (i.e. number of shapes, number of changing relations) but may vary in their distractors (two unique sets per item) or perceptual features (three unique shape sets per item). Thus, the MaRs-IB addresses several issues associated with other matrix reasoning tests.

\cite{chierchia2019matrix} also conducted an initial study of the psychometric properties of the MaRs-IB. A sample of 659 participants completed a fixed-length and fixed-order test measure in which they had eight minutes to complete as many MaRs-IB items as possible. The authors found that performance on this test measure had good split-half and test-retest reliability. Moreover they found that participants' ability to solve puzzles in the MaRs-IB was moderately predictive of their performance on a digit span task and the ICAR matrix reasoning test, indicating good convergent validity. At the conclusion of this study, \cite{chierchia2019matrix} publicly released a summary of the difficulty (i.e. proportion correct responses) of each item in the MaRs-IB so that future researchers could construct new test measures of custom difficulty and duration.

Unfortunately, there are two critical limitations in this study that undermine the utility of the summary statistics. The first is that, given design of the test, there was relatively little data collected for the majority of the items in the MaRs-IB. Indeed, only half (42 of 80 items) were completed by 100 or more participants. As such, one cannot be confident about the reported functioning of a majority of the items (an issue that is exacerbated when considering that each item has six clones). A second issue is participants were allowed to choose for themselves whether to prioritize accuracy (number of correct responses) or productivity (number of items reached). As a consequence, the response data for items that came later in the fixed-order test were increasingly likely to be produced by participants sacrificing accuracy for speed, thereby biasing the proportion correct measure for these items (see the Supplementary materials for evidence of this point). Therefore, additional investigation of the functioning and psychometric properties of the MaRs-IB is warranted.

@@

The remainder of the paper proceeds as follows. First we report a calibration study, in which a large sample of participants each completed a subset of items from the MaRs-IB. Using their response data, we then fit a series of multilevel item structure models in order to estimate item parameters for each item (and their clones). Using these same models, we also interrogate the relationship between item attributes and functioning within and cross items in the MaRs-IB. Then we report a second validation study, in which a smaller sample of participants completed one of three new MaRs-IB short-form tests. These tests --- designed using optimal test assembly methods applied to the item parameters estimated during the calibration study --- were found to have good psychometric properties and convergent validity. This second study provides a blueprint for how to construct new MaRs-IB test measures using the item parameter estimates that have been released as part of this study.

All data, code, and model outputs (including the item parameter estimtes) are publicly available at: \url{https://github.com/dawlab/mars-irt}.

@@

To ensure data quality, the data from multiple participants were excluded prior to analysis (see Exclusion Criteria below) leaving the data from N=1501 participants for analysis. In these participants, the majority identified as women (men: N=670; women: N=811; non-binary or other: N=13; rather not say: N=2) and the average age was 28.7 years old (sd = 9.9). The sample was relatively well-educated with the majority having completed a Bachelor's degree (N=507) or Master's degree or higher (N=322). Comparatively fewer participants completed only some college (N=471), only a high school degree (N=199), or preferred not to say (N=2).

@@

The presentation of each item was preceded by a fixation cross, which lasted for 1200 ms. Upon presentation of the item, participants were given up to 30 s to solve the puzzle. After 25 s elapsed, a clock appeared to count down the remaining 5 s. A trial ended when participants responded, or after 30 s had elapsed without response. Before the trials began, participants reviewed instructions and were made to correctly complete three practice items. Participants were instructed to respond carefully, but to guess if they could not solve the puzzle before the timer ran out. The format of the experiment and instructions were adapted from code publicly released as part of the original study (\url{https://app.gorilla.sc/openmaterials/36164}).

@@

@@

\item $\sum_{n=1}^N Q_{jn} \delta_{\beta n}$: the effect of the template (level 1) attributes on item difficulty, where $Q_{jn}$ is the value of attribute $n$ for item template $j$, and $\delta_{\beta n}$ is the effect of attribute $n$.

@@

In our analyses, we considered two template level (level 1) attributes: feature number, or the number of elements in each cell of an item; and rule number, or the number of changing relationships across the cells of an item (e.g. color change, shape change, position change). We predicted that both attributes would be positively associated with item difficulty; that is, all else equal, items with a great number of features or rules would be more difficult to solve. In contrast, we had no hypotheses about the relationship between these attributes and item discrimination.

@@

The AMIS model framework provides a natural means of accomplishing the three goals of the calibration the study. By fitting an AMIS model, we estimate the item parameters and quantify the psychometric properties for every item, and their clones, in the MaRs-IB (Aim 1). By modeling the effects of item attributes, we can evaluate the success of the MaRs-IB design by investigating whether item attributes influence item functioning as expected (e.g. a positive association between rule number and item difficulty; Aim 2). Finally, by estimating the magnitude of residual error in item parameters from unmodeled sources of variance (e.g. perceptual features), we can test whether the MaRs-IB possesses certain desirable structural properties (e.g. psychometric equivalence and exchangeability among item clones; Aim 3).

@@

Across all 384 item clones, the average proportion correct was 0.60 (sd = 0.19, IQR = 0.45 - 0.74; Figure \ref{fig:fig01}b). A total of 12 items (3.1\%) exhibited performance levels beneath chance, and only 22 items (5.7\%) had performance levels near ceiling ($\geq$ 90\%). Across items, the proportion of correct responses was negatively correlated with rule number ($\rho$ = -0.482, p < 0.001). These results further corroborate those reported by \cite{chierchia2019matrix} in that the MaRs-IB exhibited good functioning with performance levels for only a few item clones at ceiling or floor.

@@

We fit AMIS models in two waves. In the first wave of models, the item discrimination parameters were fixed ($\alpha = 1$).\footnote{Note that these models still included a fixed guessing parameter ($\gamma = 0.25$). We also estimated models with no guessing parameter (i.e. one-parameter logistic or Rasch models), but model comparison indicated these models provided worse fits to the data (Table SXX).} We applied this restriction to the first set of models in order to identify the best structure for the item difficulty parameters independent of the discrimination parameters. In the first wave, we fit four models:

@@

\item Model 1: Item difficulty is a function only of item template attributes (i.e. feature number, rule number). In this model, item with identical attributes are psycho- metrically equivalent, as are all the clones belonging to an item.

@@

\item Model 3: Item difficulty is a function of both item template and clone attributes (i.e. distractor type, mean response time), as well as template-level residuals. Item clones may now vary in their difficulty as a function of their attributes.

\item Model 4: Item difficulty is function of attributes and residuals at both levels. In this model, neither item templates nor clones with identical attributes are necessarily equally difficult, depending on the magnitude of the residual terms.

@@

Feature number, rule number, and distractor type cumulatievely explained approximately 39.2\% of the variance in item discrimination across the 64 item templates. In contrast to item difficulty, model comparison indicated there was non-negligible residual variability in item discrimination only at template level. Indeed, a model that included random effects at the clone level produced a comparatively worse fit to the data. As such, the results indicate that item templates with identical attributes cannot be assumed to be equally discriminating; however, there was not sufficient evidence to refute that item clones are equally discriminating. We note this result should be interpreted with caution given the relatively weak predictive improvement of the best-fitting model compared to other similar models.

@@

Next, we proceeded to administer each short form to a second sample of participants. Our aim was to verify that observed scores on each short form conformed to our expectations based on the TCCs and TIFs. Successfully doing so would provide additional evidence that we were successful in calibrating the functioning of the items in the MaRs-IB. We also sought to establish the convergent validity by correlating observed scores on the short form measures with performance on an abbreviated version of the Raven's progressive matrices.

@@

To ensure data quality, the data from multiple participants were excluded prior to analysis (see Exclusion Criteria below) leaving the data from N=300 participants for analysis. The majority of participants identified as women (men: N=138; women: N=147; non-binary or other: N=13; rather not say: N=2). Participants were 35.0 years old (sd = 13.2) on average. The sample was relatively well-educated, with the majority having completed a Bachelor's degree (N=121) or Master's degree or higher (N=45). By comparison, fewer participants endorsed having completed only some college (N=100), only a high school degree (N=33), or preferred not to say (N=1).

@@

The partial correlation between observed scores on the RPM and ability on the MaRs-IB was $\rho = 0.527$ (95\% HDI = 0.457 - 0.577). Considering the attenuation of correlations between two imperfect measures \citep{spearman1961proof}, we should not expect correlations much larger than $\rho \approx 0.64$ (or $0.80 \times 0.80$). Therefore, this result indicates good convergent validity between the two matrix reasoning measures. Briefly, in this second sample, neither age nor gender was credibly associated with performance on the MaRs-IB short forms (age: $\rho$ = -0.123, 95\% HDI = -0.253 -- 0.013; gender: $\rho$ = -0.025, 95\% HDI = -0.160 -- 0.108).

@@

\bibliography{references}

@@

\begin{table}
    \centering
    \begin{tabular}{cccccccc}
    \toprule
    Model & FE-1 & RE-1 & FE-2 & RE-2 & psis-loco & $\Delta$ psis-loco (se) \\
    \midrule
    1 & x &   &   &   & -14516.85 & -1223.69 (44.94)\\
    2 & x & x &   &   & -13804.74 & -511.58 (28.77) \\
    3 & x & x & x &   & -13481.12 & -187.96 (17.25) \\
    4 & x & x & x & x & -13293.16 & - \\
    \midrule
    5 & x &   & x &   & -13209.38 & -2.13 (2.04) \\
    6 & x & x & x &   & -13207.25 & - \\
    7 & x & x & x & x & -13208.69 & 1.44 (0.85) \\
    \bottomrule
    \end{tabular}
    \caption{\label{tab:2} \normalfont Goodness of fit of item response models to MaRs-IB data. PSIS = Pareto-smoothed importance sampling. Note: LOCO values are presented in original, not deviance, scale so larger numbers indicate better fit.}
    \label{table:2}
\end{table}

\begin{table}
    \centering
    \begin{tabular*}{\textwidth}{llcccc}
    \toprule
    Measure & Sample & Task Time (sd) & Mean Score (sd) & IQR & Reliability ($\alpha$) \\
    \midrule
    MaRs-IB SF-1 & N=103 & 174.9 s (53.3) & 7.9 (2.4) & 7 - 10 & 0.761 \\
    MaRs-IB SF-2 & N=98  & 169.0 s (53.0) & 8.2 (2.6) & 7 - 10 & 0.810 \\
    MaRs-IB SF-3 & N=99  & 178.5 s (55.7) & 7.9 (2.7) & 6 - 10 & 0.829 \\
    RPM SF-A     & N=300 & 124.6 s (40.1) & 4.5 (2.0) & 3 - 9 & 0.791 \\
    \bottomrule
    \end{tabular*}
    \caption{\label{table:3} \normalfont Summary of performance on the MaRs-IB and Raven's progressive matrices (RPM) short form (SF) measures. IQR = interquartile range.}
\end{table}

\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{figures/fig00.png}
\caption{\label{fig:fig00} Example items from the MaRs-IB. (A) A simple item containing a single feature and one relation change (i.e. only the colour changes). (B) A harder item containing three features and six relation changes. In both panels, the first option is the correct solution.}
\end{figure}

>>>>

\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{xurl}
\usepackage[colorinlistoftodos]{todonotes}

@@

\title{An item response theory analysis of the Matrix Reasoning Item Bank (MaRs-IB)}
\shorttitle{IRT analysis of MaRs-IB}
\author{Samuel Zorowitz$^1$, Gabriele Chierchia$^3$, Sarah-Jayne Blakemore$^3$, Nathaniel D. Daw$^{1,2}$}
\affiliation{$^1$Princeton Neuroscience Institute, Princeton University, USA\\$^2$Department of Psychology, Princeton University, USA\\$^3$Department of Psychology, University of Cambridge, Downing Street, Cambridge, UK}

@@

Matrix reasoning tasks are among the most widely used measures of cognitive ability in the behavioral sciences. Much of their popularity undoubtedly reflects their versatility in use. Matrix reasoning tasks are strong (albeit impure) indicators of general intelligence \citep{gignac2015raven} and working memory capacity \citep{kane2004generality, unsworth2005working}. They are predictive of important real-world outcomes such as childhood academic performance \citep{roth2015intelligence} and performance on college entrance exams \citep{frey2004scholastic, koenig2008act}. In low-stakes testing settings (i.e. where poor performance incurs marginal or no cost to the participant), matrix reasoning tasks can additionally function as measures of motivation, willingness to expend mental effort, and other aspects of personality \citep{duckworth2011role, gignac2019maximum}. In studies of psychiatric populations, performance on matrix reasoning tasks have also been used to account for general disruptions to cognitive ability where specific domains of cognition are primarily of interest \citep{gillan2016characterizing, rouault2018psychiatric, moutoussis2021decision}.

Unfortunately there are a number of impediments to using matrix reasoning tasks as part of behavioral research. One such challenge is the problem of copyright. The most prominent examples of matrix reasoning tasks, such as the WAIS and WASI matrix reasoning subtests \citep{wechsler1999wechsler, corporation2008wechsler}, are not free to use and have legal restrictions against being computerized. A second challenge is that, across all of the matrix reasoning tests in the public domain, there are relatively few unique items available. The Hagen matrices test \citep{heydasch2014hagen} and ICAR matrix reasoning test \citep{condon2014international}, for example, have only 20 and 16 items respectively. The availability of only a limited number of items raises the possibility of repeat exposure effects, which threaten the validity of these measures  \citep{ng1974applicability, bors2003effect}. This problem is exacerbated in the current age of online experiments, where multiple groups of researchers may be unintentionally administering the same test to the same participants recruited from the same online labor platforms.

Noting these challenges, \cite{knoll2016window} developed and  \cite{chierchia2019matrix} made publicly available the matrix reasoning item bank (MaRs-IB), a collection of 80 unique matrix reasoning puzzles. Each puzzle in the MaRs-IB consists of a 3x3 matrix containing abstract shapes in eight out of nine cells (Figure \ref{fig:fig00}). To solve a puzzle, participants must deduce which of four response options correctly completes the matrix based on the pattern of changing relations among the shapes across cells. The puzzles in the MaRs-IB were designed to span a considerable range of complexity, both in the number of unique shapes and changing relations in and across cells. To further increase the reusability of the MaRs-IB and minimize practice effects, each puzzle (hereafter referred to as a template item) has six clones. An item's clones share identical structural properties (i.e. number of shapes, number of changing relations) but may vary in their distractors (two unique sets per item) or perceptual features (three unique shape sets per item). Thus, the MaRs-IB addresses several issues associated with other matrix reasoning tests.

\cite{chierchia2019matrix} also conducted an initial study of the psychometric properties of the MaRs-IB. A sample of 659 adult, adolescent, and child participants completed a fixed-length and fixed-order test measure in which they had eight minutes to complete as many MaRs-IB items as possible. The authors found that performance on this test measure had good split-half and test-retest reliability. Moreover they found that participants' ability to solve puzzles in the MaRs-IB was moderately predictive of their performance on a digit span task and the ICAR matrix reasoning test, indicating good convergent validity. At the conclusion of this study, \cite{chierchia2019matrix} publicly released a summary of the difficulty (i.e. proportion correct responses) of each item in the MaRs-IB so that future researchers could construct new test measures of custom difficulty and duration.

There are two critical limitations in this study that undermine the utility of the summary statistics. The first is that, given design of the test, there was relatively little data collected for the majority of the items in the MaRs-IB. Indeed, only half (42 of 80 items) were completed by 100 or more participants. As such, one cannot be confident about the reported functioning of a majority of the items (an issue that is exacerbated when considering that each item has six clones). A second issue is participants were allowed to choose for themselves whether to prioritize accuracy (number of correct responses) or productivity (number of items reached). As a consequence, the response data for items that came later in the fixed-order test were increasingly likely to be produced by participants sacrificing accuracy for speed, thereby biasing the proportion correct measure for these items (see the Supplementary materials for evidence of this point). Therefore, additional investigation of the functioning and psychometric properties of the MaRs-IB is warranted.

@@

The remainder of the paper proceeds as follows. First we report a calibration study in which a large sample of participants each completed a subset of items from the MaRs-IB. Using their response data, we then fit a series of multilevel item structure models in order to estimate item parameters for each item (and their clones). Using these same models, we also interrogate the relationship between item attributes and functioning within and cross items in the MaRs-IB. Then we report a second validation study, in which a smaller sample of participants completed one of three new MaRs-IB short-form tests. These tests --- designed using optimal test assembly methods applied to the item parameters estimated during the calibration study --- were found to have good psychometric properties and convergent validity. This second study provides a blueprint for how to construct new MaRs-IB test measures using the item parameter estimates that have been released as part of this study.

All data, code, and model outputs (including the item parameter estimates) are publicly available at: \url{https://github.com/dawlab/mars-irt}.

@@

To ensure data quality, the data from multiple participants were excluded prior to analysis (see Exclusion Criteria below) leaving the data from N=1501 participants for analysis. In these participants, the majority identified as women (men: N=670; women: N=811; non-binary or other: N=13; rather not say: N=2) and the average age was 28.7 years old (sd = 9.9; range: 18 -- 74). The sample was relatively well-educated with the majority having completed a Bachelor's degree (N=507) or Master's degree or higher (N=322). Comparatively fewer participants completed only some college (N=471), only a high school degree (N=199), or preferred not to say (N=2).

@@

The presentation of each item was preceded by a fixation cross, which lasted for 1200 ms. Upon presentation of the item, participants were given up to 30 s to solve the puzzle. After 25 s elapsed, a clock appeared to count down the remaining 5 s. A trial ended when participants responded, or after 30 s had elapsed without response. Before the trials began, participants reviewed instructions and were made to correctly complete three practice items. Participants were instructed to respond carefully, but to guess if they could not solve the puzzle before the timer ran out. The format of the experiment and instructions were adapted from code publicly released as part of the original study (\url{https://app.gorilla.sc/openmaterials/36164}). The task was programmed in jsPsych \citep{de2015jspsych} and distributed using custom web-application software (available at \url{https://github.com/nivlab/nivturk}).

@@

\subsubsection{Statistical analysis}

@@

\item $\sum_{n=1}^N Q_{jn} \delta_{\beta n}$: the effect of the template (level 1) attributes (desribed below) on item difficulty, where $Q_{jn}$ is the value of attribute $n$ for item template $j$, and $\delta_{\beta n}$ is the effect of attribute $n$.

@@

In our analyses, we considered two template level (level 1) attributes: feature number, or the number of elements in each cell of an item; and rule number, or the number of changing relationships across the cells of an item (e.g. size change, color change, shape change, position change). The feature and rule number for each item was obtained via manual annotation. Across item templates, the modal number of features was two (range: 1 -- 4) and the modal number of rules was three (range: 1 -- 6). Feature and rule number were largely uncorrelated across items ($\rho$ = 0.176). Additional examples of items are provided in the supplementary materials. We predicted that both attributes would be positively associated with item difficulty; that is, all else equal, items with a great number of features or rules would be more difficult to solve. In contrast, we had no hypotheses about the relationship between these attributes and item discrimination.

@@

It is important to note that we deliberately chose not to include shape type as a clone attribute; that is, we elected not to directly model the effects of lower-level perceptual features on item functioning. As previously mentioned, each item template has three alternate forms that vary in their perceptual stimuli. Every item clone in the MaRs-IB is comprised of stimuli drawn one of nine stimulus sets (Figure S1). Each stimulus set is itself made up of approximately 15 unique stimuli (some elementary shapes, like squares and circles, are shared across stimulus sets). These stimuli are pseudo-randomly distributed across item clones of a given stimulus set; that is, two item clones with the same stimulus set and feature number are not guaranteed to be comprised of the same stimuli. As such, the most rigorous way of modeling perceptual features on item functioning would have been the inclusion of approximately 45 additional attributes for both item difficulty and discrimination. Rather than deal with this complexity, however, we opted to leave this unmodeled, instead to be soaked up by the residual variance term.

The AMIS model framework provides a natural means of accomplishing the three goals of the calibration the study. By fitting an AMIS model, we estimate the item parameters and quantify the psychometric properties for every item, and their clones, in the MaRs-IB (Aim 1). By modeling the effects of item attributes, we can evaluate the success of the MaRs-IB design by investigating whether item attributes influence item functioning as expected (e.g. a positive association between rule number and item difficulty; Aim 2). Finally, by estimating the magnitude of residual error in item parameters from unmodeled sources of variance (e.g. lower-level perceptual features), we can test whether the MaRs-IB possesses certain desirable structural properties (e.g. psychometric equivalence and exchangeability among item clones; Aim 3).

@@

Across all 384 item clones, the average proportion correct was 0.60 (sd = 0.19, IQR = 0.45 - 0.74; Figure \ref{fig:fig01}b). A total of 12 item clones (3.1\%) exhibited performance levels beneath chance, and only 22 item clones (5.7\%) had performance levels near ceiling ($\geq$ 90\%). Across items, the proportion of correct responses was negatively correlated with rule number ($\rho$ = -0.482, p < 0.001). These results further corroborate those reported by \cite{chierchia2019matrix} in that the MaRs-IB exhibited good functioning with performance levels for only a few item clones at ceiling or floor.

\subsubsection{Response time statistics}

@@

We fit AMIS models in two waves. In the first wave, the item discrimination parameters were held fixed ($\alpha = 1$).\footnote{These models still included a fixed guessing parameter ($\gamma = 0.25$). We also estimated models with no guessing parameter (i.e. one-parameter logistic or Rasch models), but model comparison indicated these models provided worse fits to the data (Table S1).} We applied this restriction to the first set of models in order to identify the best structure for the item difficulty parameters independent of the discrimination parameters. In the first wave, we fit three models:

@@

\item Model 1: Item difficulty is a function of item template attributes (i.e. feature number, rule number) and item clone attributes (i.e. distractor type, response time). In this model, the difficulty of an item is explained solely

item with identical attributes are psycho- metrically equivalent, as are all the clones belonging to an item.

@@

\item Model 3: Item difficulty is function of attributes and residuals at both levels. In this model, neither item templates nor clones with identical attributes are necessarily equally difficult, depending on the magnitude of the residual terms.

@@

Feature number, rule number, and distractor type cumulatively explained approximately 39.2\% of the variance in item discrimination across the 64 item templates. In contrast to item difficulty, model comparison indicated there was non-negligible residual variability in item discrimination only at template level. Indeed, a model that included random effects at the clone level produced a comparatively worse fit to the data. As such, the results indicate that item templates with identical attributes cannot be assumed to be equally discriminating; however, there was not sufficient evidence to refute that item clones are equally discriminating. We note this result should be interpreted with caution given the relatively weak predictive improvement of the best-fitting model compared to other similar models.

@@

Next, we proceeded to administer each short form to a second sample of participants. Our aim was to verify that observed scores on each short form conformed to our expectations based on the TCCs and TIFs. Doing so would provide additional evidence of whether we were successful in calibrating the functioning of the items in the MaRs-IB. We also sought to establish the convergent validity by correlating observed scores on the short form measures with performance on an abbreviated version of the Raven's progressive matrices \citep{bilker2012development}.

@@

To ensure data quality, the data from multiple participants were excluded prior to analysis (see Exclusion Criteria below) leaving the data from N=300 participants for analysis. The majority of participants identified as women (men: N=138; women: N=147; non-binary or other: N=13; rather not say: N=2). Participants were on average 35.0 years old (sd = 13.2; range: 18 -- 76). The sample was relatively well-educated, with the majority having completed a Bachelor's degree (N=121) or Master's degree or higher (N=45). By comparison, fewer participants endorsed having completed only some college (N=100), only a high school degree (N=33), or preferred not to say (N=1).

@@

The partial correlation between observed scores on the RPM and ability on the MaRs-IB was $\rho = 0.527$ (95\% HDI = 0.457 - 0.577). Considering the attenuation of correlations between two imperfect measures \citep{spearman1961proof}, we should not expect correlations much larger than $\rho \approx 0.64$ (or $0.80 \times 0.80$). Therefore, this result indicates satisfactory convergent validity between the two matrix reasoning measures. Briefly, in this second sample, neither age nor gender was credibly associated with performance on the MaRs-IB short forms (age: $\rho$ = -0.123, 95\% HDI = -0.253 -- 0.013; gender: $\rho$ = -0.025, 95\% HDI = -0.160 -- 0.108).

@@

\bibliography{main}

@@

\begin{table}
    \centering
    \begin{tabular*}{\textwidth}{clcr}
    \toprule
    Model & Description & psis-loco & $\Delta$ psis-loco (se) \\
    \midrule
    1 & TK  & 27741.4 & 1155.8 (60.1) \\
    2 & TK  & 26962.1 & 376.5 (34.5) \\
    3 & TK & 26585.5 & \multicolumn{1}{c}{-} \\
    \midrule
    4 & TK & 26419.6 & 5.12 (4.07) \\
    5 & TK & 26414.5 & \multicolumn{1}{c}{-} \\
    6 & TK & 26417.8 & 3.29 (1.84) \\
    \bottomrule
    \end{tabular*}
    \caption{\label{tab:2}\normalfont Goodness-of-fit of item response models to MaRs-IB data. LOCO values are presented in deviance scale (i.e. smaller numbers indicate better fit). Abbreviations: PSIS = Pareto-smoothed importance sampling; LOCO = leave-one-cluster-out.}
    \label{table:2}
\end{table}

\begin{table}
    \centering
    \begin{tabular*}{\textwidth}{llcccc}
    \toprule
    Measure & Sample & Task Time (sd) & Mean Score (sd) & IQR & Reliability ($\alpha$) \\
    \midrule
    MaRs-IB SF-1 & N=103 & 174.9 s (53.3) & 7.9 (2.4) & 7 - 10 & 0.761 \\
    MaRs-IB SF-2 & N=98  & 169.0 s (53.0) & 8.2 (2.6) & 7 - 10 & 0.810 \\
    MaRs-IB SF-3 & N=99  & 178.5 s (55.7) & 7.9 (2.7) & 6 - 10 & 0.829 \\
    RPM SF-A     & N=300 & 124.6 s (40.1) & 4.5 (2.0) & 3 - 9 & 0.791 \\
    \bottomrule
    \end{tabular*}
    \caption{\label{table:3}\normalfont Summary of performance on the MaRs-IB and Raven's progressive matrices (RPM) short form (SF) measures. IQR = interquartile range.}
\end{table}

\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{figures/fig00.png}
\caption{\label{fig:fig00} Example items from the MaRs-IB. (A) A simple item containing one feature per cell and two relation changes (i.e. a change of shape and its position changes across rows and columns, respectively). (B) A harder item containing three features and six relation changes (i.e. three position changes, two color changes, one shape change). (C/D) Variants of Items A and B, respectively, matched on complexity (i.e. number of features and relation changes).  In all panels, the first option is the correct response.}
\end{figure}

### 2022-05-09 manuscript/supplement.tex
checkpoint manuscript before edits

\title{Supplemental materials for ``An item response theory analysis of the Matrix Reasoning Item Bank (MaRs-IB)''}
\shorttitle{Supplemental materials for ``IRT analysis of MaRs-IB''}
\author{Samuel Zorowitz$^1$, Nathaniel D. Daw$^{1,2}$}
\affiliation{$^1$Princeton Neuroscience Institute, Princeton University, USA\\$^2$Department of Psychology, Princeton University, USA}

@@

\begin{table}
\centering
\begin{tabular*}{\textwidth}{ccccccllll}
\toprule
 & & & & & & \multicolumn{4}{c}{Spearman rank correlation} \\
\cmidrule(lr){7-10}
Measure & & Mean (SD) & & IQR & &  NFC-10 & PCF-8a & SNS & MaRs-SF \\
\midrule
NFC-10 & & 25.03 (8.27) & & 20.00 -- 31.00 & & - &  &  & \\
PCF-8a   &  & 22.17 (6.29) & & 18.75 -- 26.25 & & 0.27** &  - &  &  \\
SNS   &  & 29.07 (7.48) & & 25.00 -- 35.00 & & 0.46** &  0.29** &  - &   \\
MaRs-SF & &   8.00 (2.53) & &   6.00 -- 10.00 & & -0.04 &  0.04 &  0.14* &  - \\
\bottomrule
\end{tabular*}
\captionsetup{width=1.\textwidth}
\caption{\normalfont Correlations between performance on the MaRs-IB short forms and self-report measures. IQR = interquartile range; NFC-10 = need for cognition (10-item) scale; PCF = PROMIS cognitive functioning scale (8a); SNS = subjective numeracy scale. \\ ** p < 0.001,  * p < 0.05 (not corrected for multiple comparisons)}
\end{table}
\end{document}

>>>>

\title{Supplemental materials for ``An item response theory analysis of the Matrix Reasoning Item Bank (MaRs-IB)''}
\shorttitle{Supplemental materials for ``IRT analysis of MaRs-IB''}
\author{Samuel Zorowitz$^1$, Gabriele Chierchia$^3$, Sarah-Jayne Blakemore$^3$, Nathaniel D. Daw$^{1,2}$}
\affiliation{$^1$Princeton Neuroscience Institute, Princeton University, USA\\$^2$Department of Psychology, Princeton University, USA\\$^3$Department of Psychology, University of Cambridge, Downing Street, Cambridge, UK}

@@

\begin{table}
    \centering
    \begin{tabular*}{\textwidth}{clcr}
    \toprule
    Model & Description & psis-loco & $\Delta$ psis-loco (se) \\
    \midrule
    7 & TK  & 27741.4 & 635.3 (29.7) \\
    8 & TK  & 26962.1 & 665.6 (32.2) \\
    9 & TK & 26585.5 & 633.1 (33.9) \\
    \bottomrule
    \end{tabular*}
    \caption{\label{tab:2}\normalfont Goodness-of-fit of item response models to MaRs-IB data. LOCO values are presented in deviance scale (i.e. smaller numbers indicate better fit). Abbreviations: PSIS = Pareto-smoothed importance sampling; LOCO = leave-one-cluster-out.}
    \label{table:2}
\end{table}

\begin{table}
\centering
\begin{tabular*}{\textwidth}{ccccccllll}
\toprule
 & & & & & & \multicolumn{4}{c}{Spearman rank correlation} \\
\cmidrule(lr){7-10}
Measure & & Mean (SD) & & IQR & &  NFC-10 & PCF-8a & SNS & MaRs-SF \\
\midrule
NFC-10 & & 25.03 (8.27) & & 20.00 -- 31.00 & & - &  &  & \\
PCF-8a   &  & 22.17 (6.29) & & 18.75 -- 26.25 & & 0.27** &  - &  &  \\
SNS   &  & 29.07 (7.48) & & 25.00 -- 35.00 & & 0.46** &  0.29** &  - &   \\
MaRs-SF & &   8.00 (2.53) & &   6.00 -- 10.00 & & -0.04 &  0.04 &  0.14* &  - \\
\bottomrule
\end{tabular*}
\captionsetup{width=1.\textwidth}
\caption{\normalfont Correlations between performance on the MaRs-IB short forms and self-report measures. Abbreviations: IQR = interquartile range; NFC-10 = need for cognition (10-item) scale; PCF = PROMIS cognitive functioning scale (8a); SNS = subjective numeracy scale. \\ ** p < 0.001,  * p < 0.05 (not corrected for multiple comparisons)}
\end{table}

\begin{figure}
\centering
\includegraphics[width=1.0\textwidth]{figures/figS01.png}
\caption{\label{fig:figS01} An example item clone from each of the nine stimulus sets. In each row are three clones of one item template. Though some elementary shapes (e.g. circles, squares) are shared across stimulus sets, the majority of stimuli in each set are unique.}
\end{figure}

\end{document}
